Investigating soundscapes perception through acoustic scenes simulation.

### 2017-08-23 lafaySimulatedScenePerception.tex
jfp comments

\AuthorsI{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J.F. Petiot$^{1)}$}
\AddressI{$^{1)}$ Laboratoire des Sciences du Numérique, Nantes, France.\\
\hspace*{8pt}mathieu.lagrange@cnrs.fr\\
$^{2)}$ STMS Ircam-CNRS-UPMC Institut de Recherche et Coordination Acoustique/Musique, Paris, France}

@@

\Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process. Subjects are asked to create a full soundscape by means of a dedicated software tool, coupled with a structured sound data set. This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes. A rating experiment of the soundscape pleasantness using a 7-points Likert scale is conducted to further refine the analysis of the simulated urban acoustic scenes. Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.}
\Germanorfrenchabstract{}

@@

One of the main goal of soundscape studies is to identify which components of the soundscape influence human perception \cite{aletta2016soundscape}. For example, in order to study the link between an environment (urban sound scene) and the induced human sensation (calmness), one mean is to objectify the mental representation of a specific sound scene, a calm urban sound scene. Several means have been considered in the literature to do so.

@@

First, a subject can be asked to describe a given sound following a given perceptive scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}. The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. That being said, those stimuli respectively are or can be recorded and analysed. It thus allows the experimenter to gather a coarse grain physical description of the stimuli.

@@

Second, a subject can be asked to describe a given sound environment \cite{guastavino2006ideal, dubois2006cognitive}. A large amount of quantitative and semantic information is then collected about the representation of this type of sound environment that the subject has. Unfortunately, without any reference to sound data, this representation can hardly be characterized physically.

@@

We propose in this paper to consider the use of a soundscape simulator that the subject can use to objectify her representation of a given sound environment. We believe that the use of such a device allows us to gain the benefits of the two above cited approaches. As the subject is asked to produces audio data (the signal of the simulated scene), it allows the experimenter to study a precise modal version of the mental representation of the subject that is both characterized semantically and physically, respectively in terms of the sound sources and their levels.

@@

We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}.

For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sounds sources that compose a sound environment of interest. With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the events.

@@

To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmental scenes. Results and outcomes of a series of three experiments that foster on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:

@@

To the best of our knowledge, only Bruce~\al \cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics. A lack of diversity is nonetheless mentioned by the authors that limit the strenght of the outcomes given in this study.

@@

In our approach, the simulator developed for this study\footnote{simScene, available online \url{https://bitbucket.org/mlagrange/simscene}} gives a monophonic representation of the scene, with the added benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are both expressive and useful for analysis.

@@

The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}. The experiments  1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section \ref{sec:simulation}. Experiment 2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section \ref{sec:modification}. Conclusions and discussion about future work follow in Section \ref{sec:conclusion}. Experiment 1.a is based on a pilot study described in \cite{lafay2014new}.

%Dans le cadre de cette communication, nous ne présentons de \emph{SimScene} que les fonctionnalités pertienentes pour notre étude. Pour plus de détails, se référer à \cite{rossignol2015simscene,lafay2016JAES}.

@@

\section{The simulator}
\label{sec:simulator}

@@

\emph{Simscene} is an online digital audio tool whose first version have been developed as part of the HOULE prokect\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}. The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.

@@

\emph{Simscene} is designed as a simplified audio sequencer, with sequencing parameters specifically chosen for the generation of realistic soundscapes. To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}. A track is then created for this sound source within the simulator interface. The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sounds source. Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.

@@

In order to provide the user with a sound database that is well organized and covers as much as possible the variety of sound sources that are available in urban areas, a typology of urban sounds is first defined.

@@

The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} that study how humans discriminate different kind of urban soundscapes.

@@

The literature usually distinguish between:
\begin{itemize}
\item {a sound event}: an isolated sound, sparsely distributed whose acoustical characteristics may change with respect to time
\item {a sound texture}: an isolated sound of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
\end{itemize}

The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures only. These two types also have specific scheduling procedures during the sumulation process.

@@

According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from different sources of a sound environment \cite{bregman1994auditory}. Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}. In this case, the source is a perceptual entity as events that are not isolated are grouped into a stream.

@@

Maffiolo \cite{maffiolo_caracterisation_1999} distinguish two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events. She shows that those processes leads to two abstract cognitive category respectively termed "the event sequences" and "the amorphous sequences".

@@

The event sequences are composed of salient events that can easily be recognized such as \emph{car start} or \emph{male speech}. They arise from a descriptive analysis based on the identification of the sound sources. On contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}. They build upon a holistic analysis based on global acoustic features.

@@

Concerning sound textures, \ie~sounds that have stable characteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can decide not to abstract the physical information available using some of the summary statistics discarding lower level information \cite{nelken2013ear}.

@@

\begin{figure}[t]
        \myfloatalign
        \includegraphics[width=.8\linewidth]{gfx/ch_4/3.png}
       \caption{Hierachical organisation of the isolated sounds used ofr the simulation.}\label{fig:orgDb}
\end{figure}

@@

We term sample, a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples perceived equivalently.

@@

The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E. Rosch \cite{rosch1978cognition}. The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources is. For classes with high level of abstraction that have sub classes, its collection of samples is the union of the collections of the sub classes.

@@

Accounting for the perceptual matters detailed previously, two taxonomies are built, one for event sounds and one for texture sounds. Four levels of abstraction are considered from the more generic one (level 0), to the more specific classes (level 3), leading to a taxonomy close to the one used in \cite{Salamon14}. Only three levels of abstraction are considered for the texture sounds.

@@

Original sounds have been recorded using a shotgun microphone \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} plugged to a \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}} recorder. The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background. It also allowed us to avoid dominant sound sources while recording texture sounds by targeting distant areas with no dominant sound sources.

@@

All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie relative to Full Scale. In our case, the full scale level is of 1 Volt.

@@

By a volountary design choice, the simulation tool do not allow the user to interact and control a specific sample. Interaction is done at the track level, a track being a sequence of samples. Several parameters are available to the subject to control the track:

@@

To improve simulation quality, two parameters are also proposed:

@@

Once the typology and the set of sounds are available, an important design issue os the need of a suitable way to display the sound dataset to the user. Most browsing tools are based on keyword indexing. However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potenitally induce biases in the analysis conclusions. Indeed, the subject may select \textit{a priori} sounds referenced as belonging to \emph{park} to build a \emph{calm} soundscape.

@@

Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a two dimensional plane.

@@

Each of those classes have a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle. The subject can browse the database by listening to those prototype sounds. The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed in \cite{lafay2016JAES}.

@@

As shown on Figure~\ref{fig:simscene}, the simulation interface displays a schematic of the scene under creation. Each track is represented as an horizontal band with a temporal axis. Each sample of this track is displayed by a rectangle whose height is proportional to the amplitude of the sample. For event's tracks, the horizontal spacing between those rectangles is a function of the time delay between their onsets. For tracks of texture, a unique rectangle is displayed as this kind of sounds do not allow spacing with silence. As actual amplitude and spacing values are drawn from random variables, each time the subject changes the value of a parameter, the location and height of the rectangles are updated to reflect the changes in the sequencing of the samples. The subject can listen to the resulting sound scene at any time.

@@

\section{Experiment 1}
\label{sec:simulation}

@@

Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness. For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}. Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes: respectively ideal (i) or non ideal (ni). Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on pleasantness. For that, a second group of participants are asked to evaluate the pleasantness of each simulated scene, on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item evaluate the respective influence on the pleasantness of the various sources composing the scenes (i or ni)
\item detect the presence of \emph{outliers} or ambiguous samples. For the remainder of the study, the defined hedonic properties (i and ni) are used as references. We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of percieved pleasantness.
\end{enumerate}
\end{itemize}

Our analysis is based on the data collected by these two experiments: 1.a and 1.b.

\begin{figure}[t]
        \myfloatalign
        \includegraphics[width=\linewidth]{gfx/ch_5/5.png}
        \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b)}\label{fig:xp1_2}
\end{figure}

@@

@@

The participants have to simulate two urban sound environments of one minute each. They follow the instructions below:
\begin{itemize}
\item  first simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
\item second simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
\end{itemize}

All the participants start by desiging the ideal environment; they read the second set of instruction at the end of the first experiment. Participants are completely free of their choice concerning sounds and synthesis parameters. They must nevertheless respect the two following constraints :
\begin{itemize}
\item the listening point of view is the one of a fixed listener;
\item the soundscape must be realistic, i.e. physically plausible.
\end{itemize}
For instance, the participant can insert ten dogs in the soundscape but she cannot insert one dog barking every 10 milliseconds.
These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation user interface.

@@

Each simulation process contains two parts:
\begin{enumerate}
\item composition of the simulation: this step is split in three steps:
\begin{itemize}
  \item  select the class of sounds
	\item  give a name to the selected class of sounds
	\item  set the parameters of the tracks related to the selected class of sounds, see Section~\ref{sec:simscene_parametre}.
\end{itemize}
\item production of a free verbalization of the composed soundscape
\end{enumerate}

In addition, once the two sound scenes are done, participants are invited to
\begin{itemize}
\item  point out the sound sources that were missing for the composition
\item  give a comment about the ergonomics of the simulation environment
\item   give a comment about the ergonomics of the selection tool
\end{itemize}

@@

All the participants do the experiment on desktop computers with the same hardware and software configurations. The audio files are played in diotic conditions by headphones. During the tutorial, participants are asked to adjust the sound level in order to be comfortable. Once set, they are not allowed to modify it in the remaining of the experiment.
All the participants do the experiment at the same time. They are equally split in three identical quiet rooms. They are not allowed to talk to each other during the experiment.
Three experimenters (one in each room) are present during the whole duration of the experiment in order to assist participants with potential hardware and software issues and to answer questions.

@@

44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in downtown Nantes since 2 years or more.
Among the 44 participants, 40 succeeded, producing in the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes). the 4 other participants were excluded from the process due to a lack of respect or understanding of the instructions, or for exceeding the maximum duration of the experiment.

@@

Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes composed in Experiment 1.a (from sec. 15 to sec. 45).
The assessment is done with a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant). Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli. After the evaluation, they are free to switch to the next scene.
For each participant, sound scenes are played in a random order. The first 10 scenes allow the participants to calibrate their scores; as they are always composed of 5 ideal scenes and 5 non ideal scenes. These first 10 scenes are played back again at the end of the experiment and the last evaluations are the ones that are only taken into account in the results.
The experiment is planned to last 30 minutes. The participants do not know anything beforehand about the nature of the sound scenes.

@@

All the participants do the experiment on computers with equal hardware and software configurations. The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. All the sound scenes have been re-composed on the basis of the scores obtained in the simulation experiment (Exp. 1.a). The output sound level is the same for all the participants.
All the participants do the experiment simultaneously. They are not allowed to talk to each other during the experiment.
An experimenter is present during the whole duration of the experiment in order to assist participants and possibly answer questions.

@@

10 students (2 females - mean age 23.1, SD = 1.8) from Ecole Centrale de Nantes took part to the experiment. All are living in downtown Nantes for 2 years or more. None of them took part to the previous simulation experiment (Exp. 1.a).
All the participants succeeded in doing the experiment.

@@

\subsubsection*{Checking of the pleasantness of the synthesized scenes}

>>>>

\AuthorsI{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J-F. Petiot$^{1)}$}
\AddressI{$^{1)}$ Laboratoire des Sciences du Numérique, Nantes, France.\\
\hspace*{8pt}mathieu.lagrange@cnrs.fr\\
$^{2)}$ STMS Ircam-CNRS-UPMC Institut de Recherche et Coordination Acoustique/Musique, Paris, France}

@@

\Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process. Subjects are asked to create a full soundscape by means of a software tool dedicated to sound edition, coupled with a structured sound data set. This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes. A rating experiment of the soundscape pleasantness using a 7-points Likert scale is conducted to further refine the analysis of the simulated urban acoustic scenes. Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.}
\Germanorfrenchabstract{}

@@

One of the main goal of soundscape studies is to identify which components of the soundscape influence human perception \cite{aletta2016soundscape}. For example, in order to study the link between an environment (urban sound scene) and the induced human sensation (calmness), one mean is to objectify the mental representation of a specific sound scene, a calm urban sound scene. Several means have been considered in the literature to perform this task.

@@

First, a subject can be asked to assess a given sound environnment following a given perceptive scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}. The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. That being said, those recorded stimuli can be analysed, which allows the experimenter to gather a coarse
grain physical description of them.

@@

Second, a subject can be asked to describe a given sound environment \cite{guastavino2006ideal, dubois2006cognitive}. A large amount of quantitative and semantic information is then collected about the subject's representation of this type of sound environment. Unfortunately, without any reference to sound data, this representation cannot be characterized physically.

@@

We propose in this paper to consider the use of a soundscape simulator that the subject can manage to objectify his/her representation of a given sound environment. We believe that the use of such a device allows us to gain the benefits of the two above cited approaches. As the subject is asked to produces audio data (the signal of the simulated scene), it allows the experimenter to study a precise modal version of the mental representation of the subject that is both characterized semantically and physically, respectively in terms of the sound sources and their levels.

@@

We believe that the availability of a fine grain description of the sound stimuli is of great insterest, beceause recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}.

For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to very interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment of interest. Indeed, with such a material, not only the type of sound sources is available for study, but also the exact level and audio waveform for each source and also the structural characteristics of the scene, that is the temporal distribution of the events.

@@

To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmental scenes. Results and outcomes of a series of three experiments that build on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:

@@

To the best of our knowledge, only Bruce~\al \cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics. A lack of diversity is nonetheless mentioned by the authors that limit the outcomes of the study.

@@

In our approach, the simulator developed for this study\footnote{simScene, available online \url{http://soundthings.org/research/}} gives a monophonic representation of the scene, with the added benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are, as much as possible, expressive and useful for analysis.

@@

The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}. The three experiments,  1.a \emph{simulation}, 1.b \emph{pleasantness evaluation}, and 2 \emph{pleasantness evaluation after modification of the scenes} are respectively presented in Sections \ref{sec:simulation}, \ref{sec:evaluation}, and \ref{sec:modification}. Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.
%Dans le cadre de cette communication, nous ne présentons de \emph{SimScene} que les fonctionnalités pertienentes pour notre étude. Pour plus de détails, se référer à \cite{rossignol2015simscene,lafay2016JAES}.

@@

\section{The simulator}
\label{sec:simscene}

@@

\emph{Simscene} is an online digital audio tool whose first version has been developped as part of the HOULE prokect\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}. The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.

@@

\emph{Simscene} is designed as a simplified audio sequencer, with sequencing parameters specifically chosen for the generation of realistic soundscapes. To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}. A track is then created for this sound sources within the simulator interface. The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sound sources. Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.

@@

In order to provide the user with a sound database that is well organized and covers as much as possible the variety of sound sources that are present in urban areas, a typology of urban sounds is first defined.

@@

The chosen typology is established based on the category/classes of sounds found while reviewing 16 articles or thesis manuscripts \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} that study how humans discriminate different kind of urban soundscapes.

@@

Literature usually distinguishes between:
\begin{itemize}
\item {a sound event}: an isolated sound, sparsely distributed whose acoustical characteristics may change with respect to time
\item {a sound texture}: an isolated sound of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
\end{itemize}

The simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with classes of events only and the other with classes of textures. These two types have specific simulation procedures.

@@

According to the Auditory Scene Analysis (ASA) theory, the brain processes seperately events emerging from different sources of a sound environment \cite{bregman1994auditory}. Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}. In this case, the source is a perceptual entity as events that are not isolated are grouped into a stream.

@@

Maffiolo \cite{maffiolo_caracterisation_1999} distinguishes between two different categorization processes, that differentiate themselves with the capacity of the listener to identify sound events. She shows that those processes leads to two abstract cognitive category respectively termed "the event sequences" and "the amorphous sequences".

@@

The event sequences are composed of salient and identifiable events such as \emph{car start}, \emph{male speech}. They arise from a descriptive analysis based on the identification of the sound sources. On the contrary, the amorphous sequences are environments where distinct events can hardly be found, such as \emph{traffic hubhub}. They build upon an holistic analysis based on global acoustic features.

@@

Concerning sound textures, \ie~sounds that have stable caracteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can decide not to abstract the physical information available using some of the summary statistics discarding lower level information \cite{nelken2013ear}.

@@

\begin{figure}[t]
        \myfloatalign
        \includegraphics[trim={5cm 0 0 0},clip,width=.8\linewidth]{gfx/ch_4/3}
       \caption{Hierarchical organisation of the isolated sounds used in the simulation.}\label{fig:orgDb}
\end{figure}

@@

We term "sample", a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples perceived equivalently.

@@

The sound classes are organized hierachically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E. Rosch \cite{rosch1978cognition}. The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources. For classes with high level of abstraction with sub classes, their collection of samples is the union of the collections of the sub classes.

@@

Accounting for the perceptual matters detailed previously, two taxonomies are built, one for event sounds and one for texture sounds. Four levels of abstraction are considered from the more generic one (level 0), to the more specific classe (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}. Only three levels of abstraction are considered for the texture sounds.

@@

Original sounds have been recorded using a shotgun microphone \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} plugged to a \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}} recorder. The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background. It also allowed us to avoid dominant sound sources while recording texture sounds by aiming the microphone to distant areas with no dominant sound sources.

@@

All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie. relative to Full Scale. In our case, the full scale level is of 1 Volt.

@@

By design, the simulation tool does not allow the user to interact and control directly a specific sample. Interaction is done at the track level, a track being a sequence of samples. Several parameters are available to the subject to control the track:

@@

To ease the simulation, two parameters are also proposed:

@@

Once the typology and the set sound are available, an important design question that remains is to find a suitable way of presenting the data to the user. Most of browsing tools are based on keyword indexing. However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject. Indeed, the subject may select \textit{a priori} sound referenced belonging to \emph{park} to build a \emph{calm} soundscape.

@@

Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a two dimensional plane, each circle reprensenting a level 2 class.

@@

Each of those classes have a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle. The subject can browse the database by listening to those prototype sound. The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed in \cite{lafay2016JAES}.

@@

As shown on Figure~\ref{fig:simscene}, the simulation interface displays a schematic of the scene under creation. Each track is represented as an horizontal band with a temporal axis. Each sample of this track is displayed by a rectangle whose height is proportional to the amplitude of the sample. For event's tracks, the horizontal spacing between those rectangles is a function of the time delay between their onsets. For tracks of texture, a unique rectangle is displayed as this kind of sounds does not allow spacing with silence. As actual amplitude and spacing values are drawn from random variables, each time the subject changes the value of a parameter, the location and height of the rectangles are updated to reflect the changes in the sequencing of the samples. The subject can listen to the resulting sound scene at any time.

@@

\section{Expérience 1}
\label{sec:xp1}

\subsection{Objectif}

@@

Experiment 1 aims at using the simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness. For that, two first experiments are planned (\cf~Figure~\ref{fig:xp1_2}):
\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, participants have to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simscene}. Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
\item \emph{experiment 2.a (evaluation)}: actually, after the simulation phase, only a binary information on the emotional properties of the respective scenes is available: ideal (i) and non ideal (ni). This second experimental step aims at investigating our knowledge on pleasantness. For that, a second goup of participants is asked to evaluate the pleasantness of each simulated scene, on a semantic scale. This evaluation protocol has two goals:
\begin{enumerate}
\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene.
\item to detect the presence of \emph{outliers} or ambiguous samples. For the remainder of the study, the defined hedonic properties (i and ni) are used as ground truth. Then, we must have to be sure that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of pleasantness rating.
\end{enumerate}
\end{itemize}

Our analysis is based on the data collected by these two experiments (1.a, 1.b)

\begin{figure}[t]
        \myfloatalign
        \includegraphics[width=\linewidth]{gfx/ch_5/5}
        \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b)}\label{fig:xp1_2}
\end{figure}

@@

In order to validate the design of this experiment, a pilot study has been performed and is described in \cite{lafay2014new}.

@@

The participants have to simulate two urban sound environments of 1 minute each. They have to follow the instructions below:
\begin{itemize}
\item  1st simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
\item 2nd simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
\end{itemize}

All the participants start by working on the ideal environment; they read the second set of instruction at the end of the first experiment. Participants are completely free of their choices concerning sounds and synthesis parameters. They must nevertheless respect the two following constraints :
\begin{itemize}
\item the listening point of view is the one of a fixed listener;
\item  the soundscape must be realistic, i.e. physically plausible. For instance, the participant can insert ten dogs in the soundscape but he/she can’t insert one dog barking every 10 milliseconds.
\end{itemize}
These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation user interface.

@@

Each simulation process contains two parts :
\begin{enumerate}
\item composition of the simulation: this step is split in three steps
\begin{itemize}
  \item  select the class of sounds
	\item  give a name to the selected class of sounds
	\item  set the parameters of the tracks related to the selected class of sounds, see Section~\ref{sec:simscene_parametre}.
\end{itemize}
\item production of a free verbalization of the composed soundscape
\end{enumerate}

In addition, once the two sound scenes have been realized, participants are invited to
\begin{itemize}
\item  point out the sound sources that were missing for the composition
\item  give a comment about the ergonomics of the simulation environment
\item   give a comment about the ergonomics of the selection tool
\end{itemize}

@@

All the participants do the experiment on similar computers. The audio files are played in diotic conditions by headphones. During the tutorial, participants are asked to adjust the sound level in order to be comfortable; they can’t modify it in the following of the experiment.
All the participants do the experiment simultaneously. They are equally split in three identical quiet rooms. They can’t talk together during the experiment.
Three experimenters (one in each room) are present during the whole duration of the experiment in order to check that everything goes well and possibly to answer questions.

@@

44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in Nantes downtown since 2 years or more.
Among the 44 participants, 40 succeeded in doing the experiment, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes). 4 participants were excluded from the process for a lack of understanding of the instructions, or for exceeding the maximum duration of the experiment.

@@

Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes composed in Exp. 1.a (from sec. 15 to sec. 45).
The assessment is done in a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant). Before evaluating a scene, participants must listen to the first 2à seconds of the scene. After the evaluation, they are free to switch to the next scene.
For each participant, sound scenes are played in a random order. The first 10 scenes allow the participants to calibrate their marks; they imperatively gather 5 ideal scenes and 5 non ideal scenes. These first 10 scenes are played back again at the end of the experiment and the corresponding evaluations are the ones that are only taken into account in the results.
The experiment is planned to last 30 minutes. The participants don’t know anything beforehand about the nature of the sound scenes.

@@

All the participants do the experiment on similar computers. The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. The stimuli are the sound scenes obtained in the simulation experiment (Exp. 1.a). The output sound level is the same for all the participants.
All the participants do the experiment simultaneously. They can’t talk together during the experiment.
An experimenter is present during the whole duration of the experiment in order to check that everything goes well and possibly to answer questions.

@@

10 students (2 females - mean age 23.1, SD = 1.8) from Ecole Centrale de Nantes took part to the experiment. They are all living in Nantes since 2 years or more. None of them took part to the previous simulation experiment (Exp. 1.a).
All the participants succeeded in doing the experiment.

@@

\subsubsection*{Cheking of the pleasantness of the synthesized scenes}
